Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-02-20 brms_meeg.tex
adding preliminary results with actual MEG data

\abstract{Time-resolved electrophysiological measurements such as those
offered by magneto- or electro-encephalography (M/EEG) provide a unique
window onto neural activity underlying cognitive processes. Typically,
researchers are interested in testing whether and when such measures
differ across conditions and/or groups. The conventional approach
consists in conducting mass-univariate statistics through time followed
by some form of multiplicity correction (e.g., FDR, FWER) or
cluster-based inference. However, these cluster-based methods have an
important downside: they shift the focus of inference from the timepoint
to the cluster level, thus preventing any conclusion to be made about
the onset or offset of effects (e.g., differences across conditions or
groups). Here, we introduce a \emph{model-based} approch for analysing
M/EEG timeseries such as ERPs or timecourses of decoding performance and
their differences across conditions or groups. This approach relies on
Bayesian generalised additive multilevel models, which output the
posterior probabilility of the effect being above 0 (or above chance) at
every timestep, while naturally taking into account the temporal
dependencies and between-subject variability present in such data. Using
both simulation and actual EEG data, we show that the proposed approach
largely outperforms conventional methods in determining both the onset
and offset of M/EEG effects (e.g., ERPs difference, decoding
performance), producting more precise and more reliable estimates. We
provide an R package implementing the approach and illustrate how to
integrate it into M/EEG statistical pipelines in MNE-Python.}

@@

@@

From Dimigen \& Ehinger (\citeproc{ref-dimigen2021}{2021}): Recently,
spline regression has been applied to ERPs (e.g.,
\citeproc{ref-hendrix2017}{Hendrix et al., 2017};
\citeproc{ref-kryuchkova2012}{Kryuchkova et al., 2012})\ldots{} GAMMs
for EEG data (\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-meulman2015}{Meulman et al., 2015})\ldots{}

@@

Using Bayes factors for decoding performance
(\citeproc{ref-teichmann2022}{Teichmann, 2022})\ldots{}

@@

See for instance these tutorials
(\citeproc{ref-suxf3skuthy2017}{Sóskuthy, 2017};
\citeproc{ref-winter2016}{Winter \& Wieling, 2016}) or application to
phonetic data (\citeproc{ref-wieling2018}{Wieling, 2018}) or this
introduction (\citeproc{ref-baayen2020}{Baayen \& Linke, 2020}) or these
reference books (\citeproc{ref-hastie2017}{Hastie \& Tibshirani, 2017};
\citeproc{ref-wood2017}{Wood, 2017a})\ldots{} application to
pupillometry (\citeproc{ref-vanrij2019}{Rij et al., 2019})\ldots{}
GAMLSS for neuroimaging data (\citeproc{ref-dinga2021}{Dinga et al.,
2021})\ldots{}

@@

\[
\begin{aligned} 
y_{i} &\sim \mathrm{EF}\left(\mu_{i}, \phi\right)\\
g\left(\mu_i\right) &=  \underbrace{A_{i} + \mathbf{X}_{i} \gamma}_{\text{parametric part}} + \underbrace{\sum_{j=1}^{J} f_{j}\left(x_{ij}\right)}_{\text{non-parametric part}}
\end{aligned}
\]

where \(y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)\) denotes that
the observations \(y_{i}\) are distributed as some member of the
exponential family of distributions (e.g., Gaussian, Gamma, Beta,
Poisson) with mean \(\mu_{i}\) and scale parameter \(\phi\);
\(g(\cdot)\) is the link function, \(A\) is an offset,
\(\mathbf{X}_{i}\) is the \(i\)th row of a parametric model matrix,
\(\gamma\) is a vector of parameters for the parametric terms, \(f_{j}\)
is a smooth function of covariate \(x_{j}\). The smooth functions
\(f_{j}\) are represented in the model via penalised splines basis
expansions of the covariates, that are a weighted sum of basis
functions:

@@

@@

Now describe the Bayesian GAMM\ldots{} Proper inclusion of
varying/random effects in the model specification protects against
overly wiggly curves (\citeproc{ref-baayen2020}{Baayen \& Linke,
2020})\ldots{}

@@

Following the approach used by Sassenhagen \& Draschkow
(\citeproc{ref-sassenhagen2019}{2019}) and Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), we simulated EEG data
stemming from two conditions, one with noise only, and the other with
noise + signal. As in previous studies, the noise was generated by
superimposing 50 sinusoids at different frequencies, following an
EEG-like spectrum (see details and code in
\citeproc{ref-yeung2004}{Yeung et al., 2004}). As in Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), the signal was generated
from truncated Gaussian with an objective onset at 160 ms, a peak at 250
ms, and an offset at 342 ms. We simulated this signal for 250 timesteps
between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated such
data for a group of 20 participants with 50 trials per participant and
condition (Figure~\ref{fig-eeg}).

@@

We then fitted a Bayesian GAM using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018};
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). We used the
default priors in \texttt{brms}, that is, weakly informative priors. We
ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\). The \texttt{brms} package uses the same syntax as the
\texttt{R} package \texttt{mgcv} v 1.9-1 (\citeproc{ref-mgcv}{Wood,
2017b}) for specifying smooth effects.

@@

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

\caption{\label{fig-post-prob-ratio}Ratio of posterior probability
according to the GAM. Timesteps above threshold (10) are highlighted in
green.}

@@

Next we fit a hierarchical/multilevel GAM using summary statistics of
ERPs (mean and SD) at the participant level (similar to what is done in
meta-analysis).

@@

\CommentTok{\# fitting the GAM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

We compared the ability of the GAMM to correctly estimate the onset and
offset of the ERP difference to widely used methods. First, we conducted
mass-univariate t-tests (thus treating each timestep independently) and
identified the onset and offset of the ERP difference as the first and
last values crossing an arbitrary significance threshold
(\(\alpha = 0.01\)). We then followed the same approach but after
applying different forms of multiplicity correction the the p-values. We
compared two methods that control the false discovery rate (FDR) (i.e.,
BH95, \citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995}; and
BY01, \citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001}), one
method that controls the familywise error rate (FWER) (i.e.,
Holm--Bonferroni method, \citeproc{ref-holm1979}{Holm, 1979}), and two
cluster-based methods (permutation with a single threshold and TFCE,
\citeproc{ref-smith2009}{Smith \& Nichols, 2009}). The BH95, BY01, and
Holm corrections were applied to the p-values using the
\texttt{p.adjust()} function in \texttt{R}. The cluster-based inference
was implemented using a cluster-sum statistic of squared t-values, as
implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), called via the \texttt{R}
package \texttt{reticulate} v 1.35.0 (\citeproc{ref-reticulate}{Ushey et
al., 2024}). We also compared these estimates to the onset and offset as
estimated using the binary segmentation algorithm, as implemented in the
\texttt{R} package \texttt{changepoint} v 2.2.4
(\citeproc{ref-changepoint}{Killick et al., 2022a}), and applied
directly to the squared t-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}). For visualisation
and interpretability purposes, we converted p-values to s-values, which
can be interpreted as bits of surprising information, assuming a null
effect (\citeproc{ref-greenland2019}{Greenland, 2019})
(Figure~\ref{fig-corrections}).

@@

Assessing the reliability of the proposed approach using some sort of
split-half reliability (e.g., \citeproc{ref-rosenblatt2018}{Rosenblatt
et al., 2018})? Using the MEG decoding results of Nalborczyk et al.~(in
preparation) in 33 participants:

@@

@@

\subsection{Application to actual EEG data
(reliability)}\label{application-to-actual-eeg-data-reliability}

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

\section{Application to time-resolved decoding results (accuracy over
time)}\label{application-to-time-resolved-decoding-results-accuracy-over-time}

We conducted time-resolved multivariate pattern analysis (MVPA), also
known as decoding. As a result, we have a timecourse of decoding
accuracies (e.g., ROC AUC), bounded between 0 and 1, per participant
(Figure~\ref{fig-mne-decoding})\ldots{}

\caption{\label{fig-mne-decoding}Exemplary average timecourse of binary
decoding accuracy (ROC AUC) for each participant. Group-level average
decoding accuracy is depicted as a grey background density in each
panel.}

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-mne-decoding-1.pdf}

Now, we want to \emph{test} whether the group-level average decoding
accuracy is above chance (i.e., 0.5) at each timestep. We use a similar
GAM/GP as previously, but we replace the \(\mathrm{Normal}\) likelihood
function by a \(\mathrm{Beta}\) one to account for the bounded nature of
AUC values (between 0 and 1).

>>>>

\abstract{Time-resolved electrophysiological measurements such as those
offered by magneto- or electro-encephalography (M/EEG) provide a unique
window onto neural activity underlying cognitive processes. Typically,
researchers are interested in testing whether and when such measures
differ across conditions and/or groups. The conventional approach
consists in conducting mass-univariate statistics through time followed
by some form of multiplicity correction (e.g., FDR, FWER) or
cluster-based inference. However, while allowing efficient error-rates
control, cluster-based methods have an important downside: they shift
the focus of inference from the timepoint to the cluster level, thus
preventing any conclusion to be made about the precise onset or offset
of effects (e.g., differences across conditions or groups). Here, we
introduce a \emph{model-based} approch for analysing M/EEG timeseriesn
such as ERPs or decoding performance through time, and their differences
across conditions or groups. This approach relies on Bayesian
generalised additive multilevel models and outputs the posterior
probabilility of the effect being above 0 (or above chance) at every
timestep, while naturally taking into account the temporal dependencies
and between-subject variability present in such data. Using both
simulated and actual M/EEG data, we show that the proposed approach
largely outperforms conventional methods in estimating the onset and
offset of M/EEG effects, producing more precise and more reliable
estimates. We provide an R package implementing the approach and
illustrate how to integrate it into M/EEG statistical pipelines in
MNE-Python.}

@@

\subsection{Problem statement}\label{problem-statement}

Description of cluster-based approaches (see
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019})\ldots{}
As pointed by the original authors themselves
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007}), ``there There is
a conflict between this interest in localized effects and our choice for
a global null hypothesis: by controlling the FA {[}false alarm{]} rate
under this global null hypothesis, one cannot quantify the uncertainty
in the spatiotemporal localization of the effect''
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019})\ldots{}
In contrast, the proposed approach, based on Bayesian generalised
additive multilevel models, allows quantifying the probability of
effects at the level of timesteps, voxels, sensors (etc), while
naturally taking into account spatiotemporal dependencies.

@@

See also the rERP framework (\citeproc{ref-smith2014a}{N. J. Smith \&
Kutas, 2014a}, \citeproc{ref-smith2014b}{2014b}) and Tremblay \& Newman
(\citeproc{ref-tremblay2014}{2014})\ldots{}

From Dimigen \& Ehinger (\citeproc{ref-dimigen2021}{2021}): Recently,
spline regression has been applied to ERPs (e.g.,
\citeproc{ref-hendrix2017}{Hendrix et al., 2017};
\citeproc{ref-kryuchkova2012}{Kryuchkova et al., 2012})\ldots{} GAMMs
for EEG data (\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023})\ldots{}

@@

Recently, Teichmann (\citeproc{ref-teichmann2022}{2022}) provided a
detailed tutorial on using Bayes factors (BFs) to analyse the 1D or 2D
output from MVPA, that is, for testing, at every timestep, whether
decoding performance is above chance level. However, this approach
provides timeseries of BFs that ignores temporal dependencies..

@@

See for instance these tutorials
(\citeproc{ref-suxf3skuthy2017}{Sóskuthy, 2017};
\citeproc{ref-winter2016}{Winter \& Wieling, 2016}) or application to
phonetic data (\citeproc{ref-suxf3skuthy2021}{Sóskuthy, 2021};
\citeproc{ref-wieling2018}{Wieling, 2018}) or this introduction
(\citeproc{ref-baayen2020}{Baayen \& Linke, 2020}) or these reference
books (\citeproc{ref-hastie2017}{Hastie \& Tibshirani, 2017};
\citeproc{ref-wood2017}{Wood, 2017a})\ldots{} application to
pupillometry (\citeproc{ref-vanrij2019}{Rij et al., 2019})\ldots{}
GAMLSS for neuroimaging data (\citeproc{ref-dinga2021}{Dinga et al.,
2021})\ldots{} Modelling auto-correlation in GAMMs + EEG example
(\citeproc{ref-baayen2018}{Baayen et al., 2018})\ldots{}

@@

\[
\begin{aligned} 
y_{i} &\sim \mathrm{EF}\left(\mu_{i}, \phi\right)\\
g\left(\mu_i\right) &= \underbrace{\mathbf{A}_{i} \gamma}_{\text{parametric part}} + \underbrace{\sum_{j=1}^{J} f_{j}\left(x_{ij}\right)}_{\text{non-parametric part}}
\end{aligned}
\]

where \(y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)\) denotes that
the observations \(y_{i}\) are distributed as some member of the
exponential family of distributions (e.g., Gaussian, Gamma, Beta,
Poisson) with mean \(\mu_{i}\) and scale parameter \(\phi\);
\(g(\cdot)\) is the link function, \(\mathbf{A}_{i}\) is the \(i\)th row
of a known parametric model matrix, \(\gamma\) is a vector of parameters
for the parametric terms (to be estimated), \(f_{j}\) is a smooth
function of covariate \(x_{j}\) (to be estimated as well). The smooth
functions \(f_{j}\) are represented in the model via penalised splines
basis expansions of the covariates, that are a weighted sum of basis
functions:

@@

Terminology: \emph{splines} are functions composed of simpler functions.
These simpler functions are basis functions (e.g., a cubic polynomial)
and the set of basis functions is a \emph{basis}. Each basis function
gets its coefficient and the resultant spline is the sum of these
weighted basis functions.

@@

Now describe the Bayesian GAMM (\citeproc{ref-miller2025}{Miller,
2025})\ldots{} Proper inclusion of varying/random effects in the model
specification protects against overly wiggly curves
(\citeproc{ref-baayen2020}{Baayen \& Linke, 2020})\ldots{} Generalising
to scale and shape or ``distributional GAMs''
(\citeproc{ref-rigby2005}{Rigby \& Stasinopoulos, 2005};
\citeproc{ref-umlauf2018}{Umlauf et al., 2018}) and applied to
neuroimaging data (\citeproc{ref-dinga2021}{Dinga et al., 2021})\ldots{}

Instead of averaring, obtain the smooth ERP signal from multilevel
GAM\ldots{} less susceptible to outliers (Meulman et al., 2023)\ldots{}

@@

Following the approach of Sassenhagen \& Draschkow
(\citeproc{ref-sassenhagen2019}{2019}) and Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), we simulated EEG data
stemming from two conditions, one with noise only, and the other with
noise + signal. As in previous studies, the noise was generated by
superimposing 50 sinusoids at different frequencies, following an
EEG-like spectrum (see details and code in
\citeproc{ref-yeung2004}{Yeung et al., 2004}). As in Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), the signal was generated
from truncated Gaussian with an objective onset at 160 ms, a peak at 250
ms, and an offset at 342 ms. We simulated this signal for 250 timesteps
between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated such
data for a group of 20 participants with 50 trials per participant and
condition (Figure~\ref{fig-eeg}).

@@

We then fitted a Bayesian GAM using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018};
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). We used the
default priors in \texttt{brms} (i.e., weakly informative priors). We
ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\). The \texttt{brms} package uses the same syntax as the
\texttt{R} package \texttt{mgcv} v 1.9-1 (\citeproc{ref-mgcv}{Wood,
2017b}) for specifying smooth effects. Figure~\ref{fig-post-prob-test}
shows the predictions of this model together with the raw data.

@@

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

\caption{\label{fig-post-prob-ratio}Ratio of posterior probability
according to the GAM (on a log10 scale). Timesteps above threshold (10)
are highlighted in green.}

@@

Next we fit a hierarchical/multilevel GAM using summary statistics of
ERPs (mean and SD) at the participant level (similar to what is done in
meta-analysis). Although it is possible to fit a GAMM at the
single-trial level, we present a resource-efficient version of the model
that is fitted directly on the by-participant summary statistics (mean
and SD)\ldots{}

@@

\CommentTok{\# fitting the GAM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

We compared the ability of the GAMM to correctly estimate the onset and
offset of the ERP difference to widely used methods. First, we conducted
mass-univariate t-tests (thus treating each timestep independently) and
identified the onset and offset of the ERP difference as the first and
last values crossing an arbitrary significance threshold
(\(\alpha = 0.01\)). We then followed the same approach but after
applying different forms of multiplicity correction the the p-values. We
compared two methods that control the false discovery rate (FDR) (i.e.,
BH95, \citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995}; and
BY01, \citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001}), one
method that controls the familywise error rate (FWER) (i.e.,
Holm--Bonferroni method, \citeproc{ref-holm1979}{Holm, 1979}), and two
cluster-based methods (permutation with a single threshold and TFCE,
\citeproc{ref-smith2009}{S. Smith \& Nichols, 2009}). The BH95, BY01,
and Holm corrections were applied to the p-values using the
\texttt{p.adjust()} function in \texttt{R}. The cluster-based inference
was implemented using a cluster-sum statistic of squared t-values, as
implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), called via the \texttt{R}
package \texttt{reticulate} v 1.35.0 (\citeproc{ref-reticulate}{Ushey et
al., 2024}). We also compared these estimates to the onset and offset as
estimated using the binary segmentation algorithm, as implemented in the
\texttt{R} package \texttt{changepoint} v 2.2.4
(\citeproc{ref-changepoint}{Killick et al., 2022a}), and applied
directly to the squared t-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}). For visualisation
and interpretability purposes, we converted p-values to s-values, which
can be interpreted as bits of surprising information, assuming a null
effect (\citeproc{ref-greenland2019}{Greenland, 2019})
(Figure~\ref{fig-corrections}).

@@

Next, we assessed the performance of the proposed approach on actual MEG
data (decoding results of Nalborczyk et al., in preparation). We
conducted time-resolved multivariate pattern analysis (MVPA), also known
as decoding\ldots{} As a result, we have a timecourse of decoding
performance (ROC AUC), bounded between 0 and 1, for each participant
(for a total of 32 participants). Now, we want to \emph{test} whether
the group-level average decoding accuracy is above chance (i.e., 0.5) at
each timestep (Figure~\ref{fig-decoding-data}). We fitted a similar GAM
as discussed previously, but we replaced the \(\mathrm{Normal}\)
likelihood function by a \(\mathrm{Beta}\) one to account for the
bounded nature of AUC values (between 0 and 1) (for a tutorial on Beta
regression, see \citeproc{ref-coretta2025}{Coretta \& Bürkner, 2025}).
Note that although we chose a basis dimension of \(k=50\), which seems
appropriate for the present data, this choice should be adapted
according to the properties of the modelled data (e.g., signal-to-noise
ratio, prior low-pass filtering, sampling rate, etc) and should be
assessed by the usual model checking tools (e.g., posterior predictive
checks). We also define a region of practical equivalence (ROPE,
\citeproc{ref-kruschke2017}{Kruschke \& Liddell, 2017}), defined as the
chance level plus the standard deviation of the (group-level average)
decoding performance during the baseline period. This ensures
that\ldots{}\footnote{Could an alternative approach be to include a
  predictor for ``baseline vs.~after baseline''?}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# fitting the Beta GAM}
\NormalTok{meg\_decoding\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
\NormalTok{    auc }\SpecialCharTok{\textasciitilde{}} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{50}\NormalTok{),}
    \AttributeTok{data =}\NormalTok{ decoding\_df,}
    \AttributeTok{family =} \FunctionTok{Beta}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

We assessed the reliability of the proposed approach using a form of
permutation-based split-half reliability (e.g.,
\citeproc{ref-rosenblatt2018}{Rosenblatt et al., 2018}), which consisted
of the following steps:

@@

\caption{\label{fig-decoding-data}Group-level average decoding
performance (N=32) superimposed with the GAM predictions (in blue) and
the region of practical equivalence (ROPE, in orange) computed from the
baseline period. The blue horizontal line indicates the timesteps at
which the posterior probability ratio is equal to or greater than 20.}

\includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-decoding-data-1.pdf}

@@

\subsection{Application to actual MEG data
(reliability)}\label{application-to-actual-meg-data-reliability}

Figure~\ref{fig-onset-offset} shows the group-level average decoding
performance through time with onset and offset estimates for each
method. The \texttt{brms\_full} method is similar to the \texttt{brms}
method except that the ROPE is defined on the entire dataset rather than
on the split dataset. Overall, this figure shows that both the
\texttt{Raw\ p-value} and \texttt{FDR\ BH95} methods were extremely
lenient, considering that the decoding performance was above chance
before the onset of the stimulus (false positive) and until the end of
the trial. The \texttt{Change\ point} and \texttt{Cluster\ mass} methods
were the most conservative methods, identifying a time window from
approximately +60ms to +500ms. The \texttt{Holm}, \texttt{TFCE},
\texttt{brms}, and \texttt{brms\_full} methods produced somewhat similar
estimates of onset and offset, from approximately +60ms to
+650ms.\footnote{It should be noted that although each method can
  produce several ``clusters'' of timesteps, we only considered the
  first (onset) and last (offset) timesteps identified by each method to
  compute the error (difference).}

\caption{\label{fig-onset-offset}Group-level average decoding
performance through time with onset and offset estimates for each method
(data from Nalborczyk et al., in preparation).}

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-onset-offset-1.pdf}

Figure~\ref{fig-reliability} shows two properties of each method: i) the
median difference between the onset and offset estimates from each data
split and the onset and offset estimates from the full dataset (x-axis)
and ii) and the variance of its onset and offset estimates across data
splits (y-axis). This figure reveals that the \texttt{brms} \emph{onset
and offset} estimates on each split are the closest to the estimates
from the full dataset on average (0ms difference for the onset estimate
and 5ms difference for the offset estimate). The \texttt{Raw\ p-value}
method has similar performance, but given the aberrant estimates it
produces (cf. Figure~\ref{fig-onset-offset}), the result that it is
consistent between data splits and the full dataset is not convincing on
its own. Overall, the figure reveals that for all other methods, split
datasets produce later onset estimates and earlier offset estimates (as
compared to the estimates from the model fitted on the full dataset).

\caption{\label{fig-reliability}Median absolute error and variance of
onset (left) and offset (right) estimates for each method.}

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-reliability-1.pdf}

@@

Overall, before concluding on the onset/offset of effect based on the
model, we need to ensure that the model provides a faithful description
of the data-generating process (e.g., via posterior predictive checks
etc)\ldots{}

@@

As in previous simulation work (e.g.,
\citeproc{ref-rousselet2008}{Rousselet et al., 2008};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}), the
present simulation results depend on various choices such as the
specific cluster-forming algorithm and threshold, signal-to-noise ratio,
negative impact of preprocessing steps (e.g., low-pass filter) on
temporal resolution\ldots{} note however, that the same caveats apply to
all methods\ldots{}

@@

\bibitem[\citeproctext]{ref-baayen2018}
Baayen, R. H., Rij, J. van, Cat, C. de, \& Wood, S. (2018).
\emph{Autocorrelated errors in experimental data in the language
sciences: Some solutions offered by generalized additive mixed models}
(pp. 49--69). Springer International Publishing.
\url{https://doi.org/10.1007/978-3-319-69830-4_4}

@@

\bibitem[\citeproctext]{ref-coretta2025}
Coretta, S., \& Bürkner, P.-. C. (2025). \emph{Bayesian beta regressions
with brms in r: A tutorial for phoneticians}.
\url{http://dx.doi.org/10.31219/osf.io/f9rqg_v1}

@@

\bibitem[\citeproctext]{ref-kruschke2017}
Kruschke, J. K., \& Liddell, T. M. (2017). The Bayesian New Statistics:
Hypothesis testing, estimation, meta-analysis, and power analysis from a
Bayesian perspective. \emph{Psychonomic Bulletin \& Review},
\emph{25}(1), 178--206. \url{https://doi.org/10.3758/s13423-016-1221-4}

@@

\bibitem[\citeproctext]{ref-meulman2023}
Meulman, N., Sprenger, S. A., Schmid, M. S., \& Wieling, M. (2023).
GAM-based individual difference measures for L2 ERP studies.
\emph{Research Methods in Applied Linguistics}, \emph{2}(3), 100079.
\url{https://doi.org/10.1016/j.rmal.2023.100079}

@@

\bibitem[\citeproctext]{ref-miller2025}
Miller, D. L. (2025). Bayesian views of generalized additive modelling.
\emph{Methods in Ecology and Evolution}.
\url{https://doi.org/10.1111/2041-210x.14498}

@@

\bibitem[\citeproctext]{ref-rigby2005}
Rigby, R. A., \& Stasinopoulos, D. M. (2005). Generalized Additive
Models for Location, Scale and Shape. \emph{Journal of the Royal
Statistical Society Series C: Applied Statistics}, \emph{54}(3),
507--554. \url{https://doi.org/10.1111/j.1467-9876.2005.00510.x}

@@

\bibitem[\citeproctext]{ref-smith2014a}
Smith, N. J., \& Kutas, M. (2014a). Regression{-}based estimation of ERP
waveforms: I. The rERP framework. \emph{Psychophysiology}, \emph{52}(2),
157--168. \url{https://doi.org/10.1111/psyp.12317}

\bibitem[\citeproctext]{ref-smith2014b}
Smith, N. J., \& Kutas, M. (2014b). Regression{-}based estimation of ERP
waveforms: II. Nonlinear effects, overlap correction, and practical
considerations. \emph{Psychophysiology}, \emph{52}(2), 169--181.
\url{https://doi.org/10.1111/psyp.12320}

@@

\bibitem[\citeproctext]{ref-suxf3skuthy2021}
Sóskuthy, M. (2021). Evaluating generalised additive mixed modelling
strategies for dynamic speech analysis. \emph{Journal of Phonetics},
\emph{84}, 101017. \url{https://doi.org/10.1016/j.wocn.2020.101017}

@@

\bibitem[\citeproctext]{ref-tremblay2014}
Tremblay, A., \& Newman, A. J. (2014). Modeling nonlinear relationships
in ERP data using mixed{-}effects regression with R examples.
\emph{Psychophysiology}, \emph{52}(1), 124--139.
\url{https://doi.org/10.1111/psyp.12299}

\bibitem[\citeproctext]{ref-umlauf2018}
Umlauf, N., Klein, N., \& Zeileis, A. (2018). BAMLSS: Bayesian Additive
Models for Location, Scale, and Shape (and Beyond). \emph{Journal of
Computational and Graphical Statistics}, \emph{27}(3), 612--627.
\url{https://doi.org/10.1080/10618600.2017.1407325}

@@
